Explore the SingleR results

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
added text throughout and adjusted fig size

## Cell type comparisons

Throughout this notebook, we compare the cell type labels from `SingleR` with the consensus cell type labels.
To facilitate this, the `SingleR` labels from `NBAtlas` were grouped together to both a) mirror (where possible) the broad consensus labels, and b) to make some visualizations more manageable.

@@

@@

@@

We'll show the top-5 highest log2FC marker genes per cell type for validation.

@@

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  singler_df = full_celltype_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```

@@

@@

@@

```{r}
unique(compare_annotations_expr_df$singler) |>
  purrr::map(
    \(celltype) {

@@

>>>>

## Cell types in NBAtlas

Throughout this notebook, we compare the cell type labels from `SingleR` with the consensus cell type labels.
To facilitate this, the resulting `SingleR` labels from `NBAtlas` were grouped together to both a) mirror (where possible) the broad consensus labels, and b) to make some visualizations more manageable.

@@

We broadly see high compatibility between SingleR and consensus labels.
In addition, SingleR mostly classifies the unknown consensus cells as Neuroendocrine, which is consistent with our expectation that these are mostly tumor cells.

@@

There appears to be a reasonable correspondence between consensus- and SingleR-colored UMAPS for each cell type.
We see that SingleR classified more cells compared to consensus, which we expected, and we also see that the additional cells that SingleR labeled are roughly in the same neighborhood as their corresponding consensus labels.

@@

This dotplot shows the top-5 highest log2FC marker genes per NBAtlas cell type for validation.
Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.

This plot does not show the recoded SingleR labels but rather all labels that have associated marker genes; therefore, there are more cell type categories in this plot than in other plots in this report.

@@

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  singler_df = full_celltype_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```
For validation, we expect to see high marker gene expression along the diagonal of the dotplot where cell type labels match with their marker genes.
We do broadly see high expression along the diagonal, which means our labeled cells have some amount of similar signal to those in NBAtlas.
Some sets of marker genes show high expression across multiple cell type categories in particular for closely related cell types (e.g. monocyte, macrophage, myeloid, dendritic, or T-cells and NK-cells), but there do not appear to be any strongly unexpected gene expression patterns.

@@

This dotplot shows expression of consensus validation marker genes across matching cell types labeled with SingleR.

@@

Again, marker gene expression tends to be high in the matching cell type categories, with the possible exception of `B cell` which has somewhat lower gene expression of fewer genes; but, there is still some expected expression there.
We also see some "promiscuity" where marker genes for closely-related cell types show high expression.

@@

```{r, fig.width = 10, fig.height = 20}
unique(compare_annotations_expr_df$singler) |>
  purrr::map(
    \(celltype) {

@@

In most cases, gene expression for cells with matching consensus/SingleR annotations is higher or about the same as is gene expression for cells with different annotations.
